Supplemental Materials for ``Universals of word order reflect optimization of grammars for efficient communication''

### 2019-09-29 writeup/pnas-word-order-si.tex
code

\title{Supplemental Materials for ``Universals of word order result from optimization of grammars for efficient communication''}
\author{
        Michael Hahn \\
                Department of Linguistics\\
       Stanford University
            \and
       Daniel Jurafsky\\
       Department of Linguistics\\
       Stanford University \\
       \and
       Richard Futrell\\
       Department of Language Science\\
       University of California, Irvine
}
%\author{
%Michael Hahn, Daniel Jurafsky, Richard Futrell}

@@

The reason is that greater values of $\lambda$ can mathematically result in degenerate solutions.
To show this, note that the following inequality always holds:
\begin{equation}
\operatorname{I}[\utterance; \tree] \leq \operatorname{H}[\utterance].
\end{equation}
Therefore, if $\lambda \geq 1$, the efficiency objective satisfies 
%\begin{equation}
$R_{\textit{Eff}} = \operatorname{I}[\utterance; \tree] - \lambda \operatorname{H}[\utterance] \leq 0$,
%\end{equation}
and it takes the maximal possible value of zero if there is only a single utterance $\utterance$, in which case both $\operatorname{I}[\utterance; \tree]$ and $\operatorname{H}[\utterance]$ are zero.
This is a degenerate language with only a single utterance, which is simultaneously used to convey all meanings.
While the design of our word order grammars (see \emph{Materials and Methods}) precludes a collapse of all syntactic structures to a single utterance, this shows that an objective with $\lambda \geq 1$ cannot be a generally applicable description of the efficiency of communication systems.
In conclusion, $\lambda$ is constrained to be in $[0,1)$, with values closer to $1$ placing similar weights on both predictability and parseability, whereas values closer to $0$ diminish the role of predictability.

@@

%Finally, the robustness of our results to small variations of $\lambda$ is confime by reporting results with $\lambda=1$ in Table~\ref{table:corr-resu-previous}.

%In terms of information theory \cite{shannon1948mathematical}, this corresponds to the notion of the rate-distortion tradeoff.
%In classical rate-distortion theory, the cost of a code is its rate, the mutual information between codes and underlying messages (as in the Information Bottleneck model).
% classical models of distortion are the Hamming distortion and 
%TODO also mention rate-distortion theory

%- our function

%similar: RSA, there it's utterance-level

%- number of different forms 
%
%- complexity of the mapping 
%
%- MI between
%
%
%
%We model the first term as the 
% degree to which listeners can reconstruct dependency parse structures from an utterance, i.e., the \key{parseability} if the language. This is formalized as the amount of information that utterances $u$ provide about their underlying syntactic structures $t$:
%\begin{equation}
%	R_{Pars} := \operatorname{I}[\utterance,\tree] = \sum_{t,u} p(t,u) \log \frac{p(t|u)}{p(t)}
%\end{equation}
%where the sum runs over all possible pairs of sentences $l$ and syntactic structures $t$ in the language.
%%This quantity describes the degree to which syntactic structures can be unambiguously recovered from sentences, and quantifies how successful communication is (cf. Figure~\ref{fig:comm}).
%
%
%
%We formalize the complexity of a language as its entropy~\cite{ferreri2003least,ferrericancho2007global,futrell2017memory}.
%In our setting, this corresponds to the average word-by-word surprisal, the degree to which sentences are unpredictable from the general statistics of the language.
%Surprisal has been found to be a highly accurate and general predictor of human online processing difficulty \cite{hale2001probabilistic,levy2008expectation,smith2013effect}.
%%and can be justified  in terms of predictive coding theories of information processing in the brain \cite{friston2009predictive} as well as the general algorithmic complexity of language generation and comprehension \cite{li2008introduction}.
%
%In expectation over all utterances $u$ in a language, the negative surprisal describes the \key{predictability}, or negative entropy, of the utterances:
%\begin{equation}
%	R_{Pred} := - \operatorname{H}[\utterance] = \sum_{u} p(u) \log p(u)
%\end{equation}
%where the sum runs over all possible sentences $u$ that belong to the language.
%%In keeping with Zipf's Force of Unification, this quantity describes how homogeneous the language is, i.e., it is larger if the distribution over sentences is concentrated on a smaller number of frequent sentences. 
%
%
%
%The efficiency of a language is the weighted combination of these two scoring functions~\cite{ferreri2003least,frank2012predicting,zaslavsky2018efficient,kemp2012kinship,regier2015word,goodman2013knowledge}:
%\begin{equation}\label{eq:efficiency}
%	R_{\textit{Eff}} := R_{\textit{Parseability}} + \lambda R_\textit{Pred}
%\end{equation}
%with an interpolation weight $\lambda \in [0,1)$.
%In all results from Study 2 reported here, we set $\lambda := 0.9$ in Equation~\ref{eq:efficiency}.
%See SI appendix section 5 for mathematical discussion of $\lambda$, and robustness to other choices.
%
%

%\section{Relation to the Information Bottleneck}
%
%Our efficiency objective (Equation~\ref{eq:efficiency}), to be maximized for word order grammars $L$, is restated below.
%\begin{equation}
%    \label{eq:efficiency-again}
%    J_T(L) = \operatorname{I}[L(T); T] - \lambda \operatorname{H}[L(T)].
%\end{equation}
%This equation can be seen as a special case of the objective function proposed in \citet{tishby1999information} for general lossy compression, based on rate-distortion theory \citep{cover2006elements,harremoes2007information}. In that theory, the goal is to derive an encoding $\hat{X}=L(X)$ of some variable $X$ which preserves the information in $X$ relevant to some third variable $Y$, while minimizing irrelevant information. The three variables $Y$,$X$, and $\hat{X}$ are assumed to form a Markov chain $Y \rightarrow X \underrightarrow{L} \hat{X}$. Given those variables, the information bottleneck objective to be maximized is given in Eq.~\ref{eq:ib}.
%\begin{equation}
%    \label{eq:ib}
%    J_{X,Y}(L) = \operatorname{I}[\hat{X}; Y] - \lambda \operatorname{I}[\hat{X}; X].
%\end{equation}
%In this equation, the joint distribution of $X$ and $Y$ is seen as fixed, and the conditional distribution $L=\hat{X}|X$ is optimized.
%
%In our setting we consider a language $L$ expressing trees $T$ as utterances $U$: $T \underrightarrow{L} U$. To make our formulation more parallel to the information bottleneck, we assume the utterances $U$ are passed through a channel from producer to comprehender, and received as $\hat{U}=U$. We have the Markov chain $T \underrightarrow{L} U \rightarrow \hat{U}$, where we are assuming the conditional channel distribution $\hat{U}|U$ is fixed and the conditional distribution $L=U|T$ is to be optimized. Then our objective is parallel to Eq.~\ref{eq:ib}:
%\begin{equation}
%    \label{eq:efficiency-kinda-ib}
%    J_T(L) = \operatorname{I}[U;T] - \lambda \operatorname{I}[U;\hat{U}]. % TODO should this be \operatorname{I}[\hat{U};T]?
%\end{equation}
%Assuming the channel $U \rightarrow \hat{U}$ is noiseless, we have $U=\hat{U}$, and thus $\operatorname{I}[U;\hat{U}]=\operatorname{I}[U;U] = \operatorname{H}[U]$. Then Eq.~\ref{eq:efficiency-kinda-ib} reduces to a form of Eq.~\ref{eq:efficiency-again}:
%\begin{equation*}
%    J_T(L) = \operatorname{I}[U;T] - \lambda \operatorname{H}[U].
%\end{equation*}
%
%Our efficiency objective is thus closely related to the information bottleneck; the difference is that we are optimizing the mapping from underlying meanings to representations (strings) assuming that these strings will be transmitted in a fixed noiseless channel.
%
%

@@

\begin{table}
\centering
\small{
\begin{tabular}{l||ll|lll|lll}
Language & Pred. (t) & Parse. (t) & \multicolumn{3}{c|}{Pred. (Binomial)} & \multicolumn{3}{c}{Parseab. (Binomial)} \\ 
&  $p$ & $p$ &  Est. &CI & $p$ & Est. & CI & $p$  \\ \hline \hline
\input{../results/plane/unlexicalized/per-language-statistics-unlexicalized.tex}
\end{tabular}
}
	\caption{Per-language results in Study 1, with \emph{lexicalized} parsers and word-level-only language models. Compare Table~\ref{fig:pareto-per-lang-stats}}
\end{table}

@@

@@

Efficiency combines predictability and parseability.
In the main paper (Study 1), we showed that languages tend to be optimized for these two quantities.
In this section, we show that languages are also optimized for the \emph{combination} of the two objectives.

We quantify the degree of optimality in two ways.
In Analysis 1, we consider the efficiency metric $R_\lambda := R_{Pars} + \lambda R_{Pred}$ with the tradeoff parameter $\lambda \in [0,1)$.
For each possible value of $\lambda \in [0,1)$ trading off parseability and predictability, we quantify how many baseline grammars are less efficient than the real language.
In Analysis 2, we quantify what fraction of baseline languages \emph{Pareto-dominates} the real language, showing that at most a very small fraction of grammars are more efficient than the real language.

@@

Recall the efficiency metric $R_\lambda := R_{Pars} + \lambda R_{Pred}$ with the tradeoff parameter $\lambda \in [0,1)$.
For each $\lambda$, we can quantify the fraction of baselines that have better efficiency than the real language
\begin{equation}
	O_\lambda := \frac{\# \{b \in B: R_{\lambda}(b) > R_{\lambda}(r)\}}{\# B}
\end{equation}
where $B$ is the set of baseline languages (make figure).

@@

TODO also visualize and explain

@@

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/pareto-plane-perLanguage-isoCurve-smooth.pdf}
	\caption{Per-language results. Baseline distributions are estimated using Kernel Density estimation. Orange curves connect points estimated to be dominated by the same number of baselines as the real language, obtained by Bayesian inference using a Dirichlet Gaussian Process model.}\label{fig:isoline}
\end{figure}

@@

%We eliminate this anti-conservative bias by conducting joint Bayesian inference of $\lambda$ of $O$, using the Dirichlet Gaussian Process, a Bayesian version of Kernel Density Smoothing.
%The 50 samples represent an underlying baseline distribution in the predictability-parseability plane.
%We approximate this underlying distribution as a continuous distribution from the Dirichlet Gaussian Process, and infer its parameters using Bayesian inference.
%We use this method to obtain a Bayesian estimate of the distribution of baseline distributions in the Surprisal-Parseability plane.
%The Dirichlet Gaussian Process (CITE) defines a prior over possible baseline distributions.
%Under this prior, each possible distribution is parameterized by a number $n$ of components, and $n$ Gaussians $\mu_n, \sigma^2_n$.

%We assumed that the distribution

%Again, we use the Dirichlet Gaussian Process to obtain an estimate of the posterior of $PO$.
%We used the same procedure to also estimate the curve connecting points $x$ in the predictability-parseability that have the same value $PO(x) = PO(r)$ -- this curve represents which baselines 
%Again, due to the smoothing implied by the Dirichlet Gaussian Process model, this curve connects not observed baseline samples, but hypothetical other baselines in the predictability-parseability plane.

%- identify lambda and quantify what fraction of the baselines are less efficient at this lambda (this is essentially the same as what Zaslavsky et al 2018 do)

%- also to make sure this isn't just because lambda can be chosen, quantify what fraction of baselines Pareto-dominate the same fraction

@@

\begin{figure} 
	\begin{center}
	\includegraphics[width=0.7\textwidth]{../results/correlations/figures/pred-eff-pred-pars-families-2.pdf} % figures/pred-eff-families-2.pdf
	\end{center}
	\caption{Order of the eight correlates across 51 languages, in the real grammars (left) and predicted by optimizing for efficiency, predictability, parseability (right). Dark blue: Verb patterner \emph{precedes} object patterner (English, Arabic, ...). Light blue: Verb patterner \emph{follows} object patterner (Japanese, Hindi , ...). White cells indicate that the relation is not annotated in the dataset for the given language.}\label{fig:per-lang}
%	\caption{Order of the eight correlates across 51 languages, in the real grammars (left) and predicted by efficiency optimization (right). Dark blue: Verb patterner \emph{precedes} object patterner(English, Arabic, ...). Light blue: Verb patterner \emph{follows} object patterner (Japanese, Hindi , ...) .}\label{fig:per-lang}
\end{figure}

@@

% and obtaining $p$-values from the resulting fit: For all eight correlations, the sign of $\alpha$ was positive ($p < 0.05$).

%We obtained qualitatively equivalent results for analyses on the levels of individual relations when estimating frequentist models individually for each correlation using \texttt{lme4} \cite{bates2015fitting} and obtaining $p$-values from the resulting fit: For all eight correlations, the sign of $\alpha$ was positive ($p < 0.05$).

@@

%\begin{figure}
%    \centering
%    
%    \includegraphics[scale=.25]{../results/correlations/figures/coverage-langmod-best.pdf}
%    \includegraphics[scale=.25]{../results/correlations/figures/coverage-parse-best.pdf}
%    \includegraphics[scale=.25]{../results/correlations/figures/coverage-two-lambda09-best.pdf}
%    
%    
%    \includegraphics[scale=.25]{../results/correlations/figures/coverage-ground.pdf}
%    \includegraphics[scale=.25]{../results/correlations/figures/coverage-depl.pdf}
%	\caption{Correlations satisfied by the best word order grammars for predictability, parseability, efficiency, by the UD languages, and by the best grammars for dependency length.}
%    \label{fig:posterior-prevalence}
%\end{figure}

@@

%We also report results for grammars optimized for Dependency Length Minimization (DLM), which has been argued to be a property of efficient word order, and which we find to be predicted by efficiency optimization (see Section~\ref{sec:DLM}).
%In line with this idea, our results in Table~\ref{tab:all-predictions-1} show that optimizing for DLM makes predictions similar to efficiency optimization.
%These results suggest that it is in part through favoring short dependencies that efficiency predicts word order universals, an idea that has been proposed in prior theoretical studies \citep{hawkins1994performance, hawkins2004efficiency}.
%%Previous theoretical studies arguing for efficiency explanations of word order universals have already suggested that principles akin to DLM explain the Greenberg correlations ; our results provide the first large-scale empirical confirmation of this idea, and show that this can be reduced to efficiency optimization.
%See Section~\ref{sec:DLM} for further discussion.
%

\begin{table*} % TODO figure out away to make this table more aesthetic
	\begin{center}	
\includegraphics[width=0.8\textwidth]{si-table-perrel-1a-1.png}
\end{center}
\caption{Predictions on UD relations with predictions from the typological literature.
The first section contains the eight correlations discussed in the main paper (See Section~\ref{sec:correlations}); the second section provides other relations for which predictions are available.
The `Real' column provides the prevalence among the 51 languages in the Universal Dependencies data.
We provide posterior prevalences for grammars optimized for Efficiency, and for grammars optimized for Pars(eability) and Pred(ictability) alone, obtained from the Bayesian mixed-effects analysis controlling for languages and language families (as in Table 2 of the main paper).
In the last column, we indicate what prevalence is expected according to the typological literature. 
%We additionally report results obtained when optimizing for Dependency Length Minimization (DLM), a heuristic approximation to efficiency that has been suggested in previous work. See Section~\ref{sec:DLM} for further discussion of DLM and its relation to efficiency. 
}
\label{tab:all-predictions-1}
\end{table*}

%
%%posterior_perRelation_Real_conj.pdf         posterior_perRelation_Real_goeswith.pdf     posterior_perRelation_Real_nummod.pdf       
%%user@user-G751JY:~/grammar-optim/results/correlations$ evince figures/posteriors/posterior_perRelation_Real_csubj.pdf 
%%user@user-G751JY:~/grammar-optim/results/correlations$ 
%
%\begin{table*} % TODO figure out away to make this table more aesthetic
%	\begin{center}	
%\small{
%\begin{tabular}{l|l|l|l|ll|l|l|llllllllllllllllllllllllllll}
%	\hline
%&	Relation & Real & DLM & Pred & Pars & Efficiency & Expected Prevalence  \\ \hline
%% From Dryer	
%\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {1}}}&	lifted\_case  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_lifted_case.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_lifted_case.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_lifted_case.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_lifted_case.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_lifted_case.pdf} & $> 50 \%$ \citep{dryer1992greenbergian}  \\
%\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {2}}}&	lifted\_cop  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_lifted_cop.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_lifted_cop.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_lifted_cop.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_lifted_cop.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_lifted_cop.pdf} & $> 50 \%$ \citep{dryer1992greenbergian}  \\
%\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {3}}}&	aux\footnotemark  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_aux.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_aux.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_aux.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_aux.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_aux.pdf}  & $<50\%$ \citep{dryer1992greenbergian} \\
%\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {4}}}&	nmod  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_nmod.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_nmod.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_nmod.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_nmod.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_nmod.pdf}  & $>50\%$ \citep{dryer1992greenbergian} \\
%\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {5}}}&	acl  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_acl.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_acl.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_acl.pdf}  &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_acl.pdf}  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_acl.pdf}   & $>50\%$  \citep{dryer1992greenbergian} \\
%\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {6}}}&	lifted\_mark  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_lifted_mark.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_lifted_mark.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_lifted_mark.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_lifted_mark.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_lifted_mark.pdf} & $> 50 \%$ \citep{dryer1992greenbergian}  \\
%\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {7}}}&	obl  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_obl.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_obl.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_obl.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_obl.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_obl.pdf}  & $>50\%$ \citep{dryer1992greenbergian} \\
%\raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {8}}}&	xcomp  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_xcomp.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_xcomp.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_xcomp.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_xcomp.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_xcomp.pdf}  & $>50\%$ \citep{dryer1992greenbergian} \\
%%\hdashline
%
%\hline
%% Other universals, not in Dryer
%&advcl  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_advcl.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_advcl.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_advcl.pdf}  &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_advcl.pdf}  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_advcl.pdf}   & $>50\%$ \citep{greenberg1963universals,diessel2001ordering} \\
%&ccomp  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_ccomp.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_ccomp.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_ccomp.pdf}  &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_ccomp.pdf}  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_ccomp.pdf}   & $>50\%$ (cf. \cite{dryer1980positional}) \\ % TODO According to Dryer 1980, there is a correlation
%&csubj  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_csubj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_csubj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_csubj.pdf}  &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_csubj.pdf}  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_csubj.pdf}   & $>50\%$ (cf. \cite{dryer1980positional}) \\% TODO According to Dryer 1980, there is a correlation
%&nsubj  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_nsubj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_nsubj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_nsubj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_nsubj.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_nsubj.pdf}  & See Section~\ref{sec:correlations} \\
%&amod  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_amod.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_amod.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_amod.pdf}  &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_amod.pdf}  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_amod.pdf}   &  $\approx 50\%$ \citep{dryer1992greenbergian} \\
%&nummod  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Real_nummod.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_DependencyLength_nummod.pdf}   &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Predictability_nummod.pdf}  &   \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Parseability_nummod.pdf}  &  \includegraphics[width=0.06\textwidth]{../results/correlations/figures/posteriors/posterior_perRelation_Efficiency_nummod.pdf}   &$\approx 50\%$ \citep[][89A, 83A]{wals} \\ 
% \hline
%%comp. &
%%\multirow{3}{*}{NP}&
%%	AP &
%\end{tabular}
%}
%
%\end{center}
%\caption{Predictions on UD relations with predictions from the typological literature. The first section contains the eight correlations discussed in the main paper (See Section~\ref{sec:correlations}); the second section provides other relations for which predictions are available. In the last column, we indicate what direction would be expected typologically.}
%\label{tab:all-predictions-1}
%\end{table*}
%

%\footnotetext{Note that we report an \emph{anti-correlation}. The \emph{aux} syntactic relation in UD has the auxiliary (verb-patterner) as its dependent, and has direction \emph{opposite} to the auxiliary-verb relation \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {3}}}. Therefore, this relation is \emph{anti-correlated} with the verb-object relation, while \raisebox{.5pt}{\textcircled{\raisebox{-.9pt} {3}}} is \emph{correlated}.  }

>>>>

\title{Supplemental Materials for ``Universals of word order reflect optimization of grammars for efficient communication''}
\author{
        Michael Hahn \\
                Department of Linguistics\\
       Stanford University
            \and
       Daniel Jurafsky\\
       Department of Linguistics\\
       Stanford University \\
       \and
       Richard Futrell\\
       Department of Language Science\\
       University of California, Irvine
}
%\author{
%Michael Hahn, Daniel Jurafsky, Richard Futrell}

@@

The reason is that greater values of $\lambda$ can mathematically result in degenerate solutions.
To show this, note that the following inequality always holds:
\begin{equation}
\operatorname{I}[\utterance; \tree] \leq \operatorname{H}[\utterance].
\end{equation}
Therefore, if $\lambda \geq 1$, the efficiency objective satisfies 
$R_{\textit{Eff}} = \operatorname{I}[\utterance; \tree] - \lambda \operatorname{H}[\utterance] \leq 0$,
and it takes the maximal possible value of zero if there is only a single utterance $\utterance$, in which case both $\operatorname{I}[\utterance; \tree]$ and $\operatorname{H}[\utterance]$ are zero.
This is a degenerate language with only a single utterance, which is simultaneously used to convey all meanings.
While the design of our word order grammars (see \emph{Materials and Methods}) precludes a collapse of all syntactic structures to a single utterance, this shows that an objective with $\lambda \geq 1$ cannot be a generally applicable description of the efficiency of communication systems.
In conclusion, $\lambda$ is constrained to be in $[0,1)$, with values closer to $1$ placing similar weights on both predictability and parseability, whereas values closer to $0$ diminish the role of predictability.

@@

@@

@@

\begin{figure}
\centering
		\includegraphics[width=0.5\textwidth]{lambda-explanation.png}

	\caption{TODO For $\lambda=0$ and $\lambda=0.9$, we provide lines connecting points that have the same efficiency value as the real language (red cross). Points to the bottom right of the lines have lower efficiency under the respective values of $\lambda$. In this example, all baseline samples (green dots) have lower efficiency than the real language, even though the real language does not outperform every baselines on both components of efficiency.}\label{fig:opt-lambda-explanation}
\end{figure}

@@

In the main paper (Study 1), we showed that languages tend to be optimized for parseability and/or predictability.
Efficiency is a combination of both components; in this section we address the question whether languages are generally optimized for efficiency, i.e., the multi-objective optimization problem of optimizing for parseability and predictability.
%Note that the numerical value of efficiency depends on the parameter $\lambda \in [0,1)$.

We quantify degree of optimality in two ways.
In Analysis 1, we consider the efficiency metric $R_\lambda := R_{Pars} + \lambda R_{Pred}$ with the tradeoff parameter $\lambda \in [0,1)$.
For each possible value of $\lambda \in [0,1)$ trading off parseability and predictability, we quantify how many baseline grammars are less efficient than the real language.
In Analysis 2, we quantify what fraction of baseline languages \emph{Pareto-dominates} the real language, showing that -- while not all real languages may be strongly optimized for both objectives individually -- at most a very small fraction of baseline grammars improve on the real language on both components.

@@

Recall the efficiency metric
\begin{equation}
	R_\lambda := R_{Pars} + \lambda R_{Pred}
\end{equation}
 with the tradeoff parameter $\lambda \in [0,1)$.
For each $\lambda$, we quantify the fraction of baselines that have better efficiency than the real language:
\begin{equation}
	O_\lambda := \frac{\# \{b \in B: R_{\lambda}(b) > R_{\lambda}(r)\}}{\# B}
\end{equation}
where $B$ is the set of baseline languages.
We illustrate this in Figure~\ref{fig:opt-lambda-explanation}.

@@

For all languages, there are some values of $\lambda$ where the real language improves at least on more than the median of the baseline grammars, i.e., $O_\lambda >0$.
In about 40 of the languages, $O_\lambda$ is even close to one across all values of $\lambda$.

This shows that, while many languages do not improve over \emph{all} baselines on \emph{both} individual components, they mostly improve over the large majority of baselines on the combined objective of efficiency, even across different values of $\lambda$.

@@

TODO maybe we can scrap this part as it is harder to explain well, if Analysis 1 sufficient.

@@

\begin{figure}
\centering
\includegraphics[width=\textwidth]{../results/plane/pareto-plane-perLanguage-isoCurve-smooth-interval.pdf}
	\caption{Real languages, even those that are not at the estimated Pareto frontier, are Pareto-dominated by very few other hypothetical grammars. Green curves: Baseline distribution, estimated using Kernel Density estimation. Black numbers indicate the fraction of baselines dominating the real language, with a 95\% credible interval, estimated using a nonparametric Bayesian method (see text). Orange curves connect points estimated to be dominated by the same fraction of baselines as the real language, estimated using the same Bayesian method (see text). }\label{fig:isoline}
\end{figure}

@@

@@

\begin{figure} 
	\begin{center}
	\includegraphics[width=0.7\textwidth]{../results/correlations/figures/pred-eff-pred-pars-families-2.pdf} 
	\end{center}
	\caption{Order of the eight correlates across 51 languages, in the real grammars (left) and predicted by optimizing for efficiency, predictability, parseability (right). Dark blue: Verb patterner \emph{precedes} object patterner (English, Arabic, ...). Light blue: Verb patterner \emph{follows} object patterner (Japanese, Hindi , ...). White cells indicate that the relation is not annotated in the dataset for the given language.}\label{fig:per-lang}
\end{figure}

@@

@@

@@

\begin{table*} 
	\begin{center}	
\includegraphics[width=0.8\textwidth]{si-table-perrel-1a-1.png}
\end{center}
\caption{Predictions on UD relations with predictions from the typological literature.
The first section contains the eight correlations discussed in the main paper (See Section~\ref{sec:correlations}); the second section provides other relations for which predictions are available.
The `Real' column provides the prevalence among the 51 languages in the Universal Dependencies data.
We provide posterior prevalences for grammars optimized for Efficiency, and for grammars optimized for Pars(eability) and Pred(ictability) alone, obtained from the Bayesian mixed-effects analysis controlling for languages and language families (as in Table 2 of the main paper).
In the last column, we indicate what prevalence is expected according to the typological literature. 
}
\label{tab:all-predictions-1}
\end{table*}

\begin{table*} 
	\begin{center}	
\includegraphics[width=0.7\textwidth]{si-table-perrel-2a-1.png}  
\end{center}
\caption{Predictions on UD relations for which no predictions are available in the typological literature.  ``Uninterpretable'' UD relations are those which collapse so many different linguistic relationships that they are not linguistically meaningful. ``UD artifact'' relations are those whose order is determined strictly by UD parsing standards, such that their order is not linguistically meaningful: these include dependencies such as the connection between two parts of a word that have been separated by whitespace inserted as a typo (\emph{goeswith}).
We provide results for grammars optimized for Efficiency, and for grammars optimized for Pars(eability) and Pred(ictability) alone.
}
\label{tab:all-predictions-2}
\end{table*}
